\section{\ttt{string}, \ttt{vector} and Iterators}

\subsection{\ttt{std::string}}

\begin{frame}[fragile]{Construction}
    \begin{center}
        \begin{tabular}{|ll|}
            \hline
            \begin{cpp}
string s1;
            \end{cpp} & Default initialization. \ttt{s1} contains \ttt{""}.\\
            \begin{cpp}
string s2(s1);
            \end{cpp} & Copy initialization. \ttt{s2} is a copy of \ttt{s1}.\\
            \begin{cpp}
string s2 = s1;
            \end{cpp} & Equivalent to \ttt{string s2(s1)}.\\
            \begin{cpp}
string s3("hello");
            \end{cpp} & \ttt{s3} is a copy of the string literal.\\
            \begin{cpp}
string s3 = "hello";
            \end{cpp} & Equivalent to \ttt{string s3("hello")}.\\
            \begin{cpp}
string s4(n, c);
            \end{cpp} & Initialize \ttt{s4} with \ttt{n} copies of a \bluett{char }\ttt{c}.\\
            \hline
        \end{tabular}
    \end{center}
    \begin{itemize}
        \item A string object is \textbf{NOT} null-terminated!
    \end{itemize}
\end{frame}

\begin{frame}[fragile]{IO}
    \begin{center}
        \begin{tabular}{|ll|}
            \hline
            \begin{cpp}
os << s
            \end{cpp} & Writes \ttt{s} onto output stream \ttt{os}.\\
            \begin{cpp}
is >> s
            \end{cpp} & Reads a string from \ttt{is} into \ttt{s}.\\
            \begin{cpp}
getline(is, s)
            \end{cpp} & Reads a line of input from \ttt{is} into \ttt{s}.\\
            \hline
        \end{tabular}
    \end{center}
    \begin{itemize}
        \item \ttt{is >> s} starts reading \textbf{from the first non-whitespace character}, and reads until the next whitespace character. The whitespace in the end \textbf{is not read and still in the stream}.
        \item \ttt{getline(is, s)} starts reading \textbf{from the next character} and reads until the first newline character. The newline character is \textbf{read, but not stored into \ttt{s}, and thrown away}.
    \end{itemize}
\end{frame}

\begin{frame}[fragile]{Operations}
    \begin{center}
        \begin{tabular}{|ll|}
            \hline
            \begin{cpp}
s.empty()
            \end{cpp} & \footnotesize Returns \bluett{true} iff \ttt{s} is empty (\ttt{""}).\\
            \begin{cpp}
s.size()
            \end{cpp} & \footnotesize Returns the number of characters in \ttt{s}.\\
            \begin{cpp}
s[n]
            \end{cpp} & \footnotesize Returns a \textbf{reference} to the character indexed \ttt{n} in \ttt{s}.\\
            \begin{cpp}
s1 + s2
            \end{cpp} & \footnotesize Returns a \ttt{string} that is the concatenation of \ttt{s1} and \ttt{s2}.\\
            \begin{cpp}
s1 = s2
            \end{cpp} & \footnotesize Copy-assignment. Replaces the content in \ttt{s1} with a \textbf{copy} of \ttt{s2}.\\
            \begin{cpp}
s1 += s2
            \end{cpp} & \footnotesize Equivalent \red{(?)} to \ttt{s1 = s1 + s2}.\\
            \begin{cpp}
==,!=
            \end{cpp} & \footnotesize Equality and inequality.\\
            \begin{cpp}
<,<=,>,>=
            \end{cpp} & \footnotesize Lexicographical-order comparison.\\
            \hline
        \end{tabular}
    \end{center}
\end{frame}

\begin{frame}[fragile]{Operations}
    Concatenation: \ttt{s1 + s2}.
    \begin{itemize}
        \item \ttt{s1} and \ttt{s2} can be a C-style string (\const\bluett{char }\ttt{*}) or a \bluett{char}.
        \item At least one of \ttt{s1} and \ttt{s2} should be \ttt{std::string}!
        \item \ttt{s + "a" + "b"} compiles, while \ttt{"a" + "b" + s} does not compile. Why?
    \end{itemize}
\end{frame}

\begin{frame}[fragile]{Operations}
    \ttt{s1 += s2} and \ttt{s1 = s1 + s2} are \textbf{NOT} actually equivalent:
    \begin{cpp}
auto n = 1000000;
std::string s = "";
for (auto i = 0; i != n; ++i)
  s += "a";
s = "";
for (auto i = 0; i != n; ++i)
  s = s + "a";
    \end{cpp}
    The first loop takes \(O(n)\) time, while the second takes \(O\left(n^2\right)\)!
    \begin{itemize}
        \item Always prefer compound assignment operators.
    \end{itemize}
\end{frame}

\begin{frame}[fragile]{Operations}
    \ttt{s.size()} returns a value of the type \ttt{std::string::size\_type}.
    \begin{itemize}
        \item An unsigned integer type.
        \item It is guaranteed to be able to store the length of any string.
        \item It is highly possible that it is \ttt{std::size\_t}, but not guaranteed.
    \end{itemize}
    \pause
    \ttt{s.length()} is equivalent to \ttt{s.size()}, but \ttt{s.size()} is preferred. \red{(Why?)}
\end{frame}

\begin{frame}[fragile]{Examples}
    Example: Convert a nonnegative integer to a string. Add leading zeros.
    \begin{cpp}
inline std::string convert(int x) {
  auto s = std::to_string(x);
  return std::string(9 - s.size(), '0') + s;
}
    \end{cpp}
    \pause
    Example: Count the number of upper-case letters, and convert them to lower-case.
    \begin{cpp}
int upper_cnt = 0;
for (decltype(s.size()) i = 0; i != s.size(); ++i)
  if (std::isupper(s[i])) {
    ++upper_cnt;
    s[i] = std::tolower(s[i]);
  }
    \end{cpp}
\end{frame}

\begin{frame}[fragile]{Range-based \bluett{for} Loops}
    Count the number of upper-case letters:
    \begin{cpp}
int upper_cnt = 0;
for (char c : s)
  if (std::isupper(c))
    ++upper_cnt;
    \end{cpp}
    \pause
    Convert upper-case letters to lower:
    \begin{cpp}
for (char &c : s)       // '&' is necessary!
  c = std::tolower(c);
    \end{cpp}
    \pause
    \begin{itemize}
        \item It is common to use \bluett{auto} in range-\bluett{for}.
        \item Looks like \ttt{Python} \bluett{for} loops?
    \end{itemize}
\end{frame}

\begin{frame}[fragile]{Example}
    Suppose \ttt{s} contains an English sentence. Convert every letter of the first word into upper-case.
    \pause
    \begin{cpp}
for (decltype(s.size()) i = 0;
    i != s.size() && !std::isspace(s[i]); ++i)
  s[i] = std::toupper(s[i]);
    \end{cpp}
    \pause
    Range-\bluett{for}:
    \begin{cpp}
for (auto &c : s) {
  if (std::isspace(c))
    break;
  c = std::toupper(c);
}
    \end{cpp}
\end{frame}