\section{Constant Expressions}

\subsection{\ttt{constexpr} Variables}

\begin{frame}[fragile]{Constant Expression}
    \begin{cpp}
// const variable initialized with constant expression is constant expression.
const int maxn = 1000;
std::array<int, maxn> arr; // OK.
    \end{cpp}
    \pause
    The \constexpr keyword:
    \begin{cpp}
constexpr int maxn = 1000; // Undoubtedly constant expression.
std::array<int, maxn> arr;
    \end{cpp}
\end{frame}

\begin{frame}[fragile]{Constant Expression}
    \begin{itemize}
        \item \constexpr explicitly declares that a variable must have a compile-time known value.
        \item \constexpr tells the compiler: You can work out its value!
        \pause
        \item But if the compiler cannot, it reports an error.
    \end{itemize}
    \begin{cpp}
int n; std::cin >> n;
constexpr int m = n; // Error.
    \end{cpp}
\end{frame}

\subsection{\ttt{constexpr} Functions}

\begin{frame}[fragile]{\ttt{constexpr} Functions}
    Since C++11, functions can be defined as \bluett{constexpr}, but with some constraints.
    \begin{cpp}
constexpr int add(int a, int b) {
  return a + b; // OK.
}
int x = add(42, 35); // run the function at compile-time!
int a, b; std::cin >> a >> b;
int y = add(a, b); // run the function at runtime.
    \end{cpp}
\end{frame}

\begin{frame}[fragile]{\ttt{constexpr} Functions}
    Parameter types and return-type of \constexpr functions must be a literal type.
    \begin{cpp}
constexpr std::string cat(const std::string &a, // Error!
                          const std::string &b) {
  return a + b;
}
    \end{cpp}
\end{frame}

\begin{frame}[fragile]{\ttt{constexpr} Functions}
    C++11 \constexpr functions must contain only a \bluett{return} statement, but C++14 allows more control-flow statements.
    \begin{cpp}
// Allowed since C++14
constexpr int power(int x, int n) {
  int result = 1;
  while (n--)
    result *= x;
  return result;
}
// The version for C++11
constexpr int power(int x, int n) {
  return n == 0 ? 1 : power(x, n - 1);
}
    \end{cpp}
\end{frame}

\begin{frame}[fragile]{More at Compile-time...}
    C++ has been trying to do more work at compile-time...
    \begin{itemize}
        \item Constructors may be \constexpr.
        \item STL containers like \ttt{std::vector}, \ttt{std::string} have \constexpr constructors and member functions since C++20.
        \pause
        \item Non-type template parameters must be integral type (integers, \bluett{bool}, \bluett{char}) before C++20...
        \begin{itemize}
            \item Floating-point types or class types are not allowed.
        \end{itemize}
        \pause
        \item Since C++20, anything can be used as template parameters.
    \end{itemize}
\end{frame}